Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-29 paper/main.tex
input changes from ContextLab#102

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the notebook's namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
user's Python environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the notebook for immediate use.

Importantly, packages installed by Davos are made available for use in the
notebook without affecting the user's Python environment or existing packages.
By default, \texttt{smuggle} statements will install missing packages (and any
missing dependencies of those packages) into a notebook-specific virtual
environment-like directory called a ``project'' (see
Sec.~\ref{subsec:projects}). In turn, \texttt{smuggle} statements executed in a
particular notebook will preferentially load packages from that notebook's
project directory whenever they are available, rather than searching for them
in the user's main Python environment. In this way, \texttt{smuggle}
statements can be substituted for \texttt{import} statements to automatically
ensure that all packages needed to run a notebook are installed and available
at runtime each time the notebook is run, without risking interfering with
dependencies of the user's other Python programs, or other Davos-enhanced
notebooks.

@@

Standard approaches to installing packages from within a notebook can alter the local Python environment in potentially unexpected and undesired ways.
For example, running a notebook that installs its dependencies via system shell commands (prefixed with ``\texttt{!}'') or IPython magic commands (prefixed with ``\texttt{\%}'') may cause other existing packages in the user's environment to be uninstalled or replaced with alternate versions.
This can lead to incompatibilities between installed packages, affect the behavior of the user's other scripts or notebooks, or even interfere with system applications.

To prevent Davos-enhanced notebooks from having unwanted side effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using custom virtual environment-like systems called ``projects.''
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use.
Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task.
However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and they function to \textit{extend} the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already).
When \texttt{smuggle} statements within that notebook are executed, any packages (or specific versions of packages) that are not already available in the user's Python environment are installed into the notebook's project directory (along with any missing dependencies of those packages).
During each \texttt{smuggle} statement's execution, Davos also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over those in the runtime environment.

@@

Next, in lines 8--9, we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 8 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment (since there is no chance the module will need to be installed).
Line 9 then loads the \texttt{joblib} package~\cite{Varo10},
installing it into the notebook's project directory if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring a particular exact version
would likely be unnecessarily restrictive. However, it is possible a
\textit{future} release could introduce some breaking change.  The
onion comment in line 9 helps ensure that the analysis notebook will continue
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
It is worth noting, however, that beyond illustrative purposes, the benefit of specifying only a maximum version for \texttt{joblib} rather than an exact version is relatively minor.
The main advantage to relaxing a version constraint in an onion comment (when a package's behavior does not differ meaningfully between versions) is that doing so increases the likelihood that a satisfactory version will already be available in the user's Python environment, and therefore Davos will not need to install a new copy in the notebook's project directory.
For large packages, this can be a worthwhile consideration; however \texttt{joblib} is very lightweight---less than 0.5 MB pre-built, with no other dependencies.
Thus a more conservative approach that guarantees an exact version is used would also be reasonable in this case.

@@

While Davos enables developers to conveniently specify all project
dependencies, there are some edge cases and limitations that are worth
considering. Many Python packages include (in their setup options) additional
dependencies that often carry their own version specifications. Although Davos
will check that the correct version of the requested top-level package is
installed and imported into the workspace, the version numbers of any
\textit{dependencies} of the requested package are not checked. In principle,
this could lead to unexpected behavior, for example if a given package's
dependencies (or dependencies of those dependencies, etc.) were left
under-specified. A developer could mitigate this by explicitly smuggling exact
version numbers of every project dependency (e.g., obtained via pip freeze).
However, for projects where the versions of dependencies of smuggled packages
also need to be precisely controlled, a lockfile or a \texttt{requirements.txt}
file produced by \texttt{pip freeze} (i.e., explicitly specifying \textit{all} packages'
version numbers) may provide a more comprehensive alternative to Davos.

>>>>

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the notebook's namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
user's Python environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the notebook for immediate use.

Importantly, packages installed by Davos are made available for use in the
notebook without affecting the user's Python environment or existing packages.
By default, \texttt{smuggle} statements will install missing packages (and any
missing dependencies of those packages) into a notebook-specific, virtual
environment-like directory called a ``project'' (see
Sec.~\ref{subsec:projects}). In turn, \texttt{smuggle} statements executed in a
particular notebook will preferentially load packages from that notebook's
project directory whenever they are available, rather than searching for them
in the user's main Python environment. In this way, \texttt{smuggle}
statements can be substituted for \texttt{import} statements to automatically
ensure that all packages needed to run a notebook are installed and available
at runtime each time the notebook is run, without risking interfering with
dependencies of the user's other Python programs, or other Davos-enhanced
notebooks.

@@

Standard approaches to installing packages from within a notebook can alter the local Python environment in potentially unexpected and undesired ways.
For example, running a notebook that installs its dependencies via system shell commands (prefixed with ``\texttt{!}'') or IPython magic commands (prefixed with ``\texttt{\%}'') may cause other existing packages in the user's environment to be replaced with alternate versions.
This can lead to incompatibilities between installed packages, affect the behavior of the user's other scripts or notebooks, or even interfere with system applications.

To prevent Davos-enhanced notebooks from having unwanted side effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.''
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use.
Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task.
However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and they function to \textit{extend} the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already).
When \texttt{smuggle} statements within that notebook are executed, any packages (or specific versions of packages) that are not already available in the user's Python environment are installed into the notebook's project directory (along with any missing dependencies of those packages).
During each \texttt{smuggle} statement's execution, Davos also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over those in the user's main environment.

@@

Next, in lines 8--9, we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 8 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment (since there is no chance the module will need to be installed).
Line 9 then loads the \texttt{joblib} package~\cite{Varo10},
installing it into the notebook's project directory if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring a particular exact version
would likely be unnecessarily restrictive. However, it is possible a
\textit{future} release could introduce some breaking change.  The
onion comment in line 9 helps ensure that the analysis notebook will continue
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
It is worth noting, however, that beyond illustrative purposes, the benefit of specifying only a maximum version for \texttt{joblib} rather than an exact version is relatively minor.
The main advantage to relaxing a version constraint in an onion comment (when a package's behavior does not differ meaningfully between versions) is that doing so increases the likelihood that a satisfactory version will already be available in the user's Python environment, and therefore Davos will not need to install a new copy in the notebook's project directory.
For large packages, this can be a worthwhile consideration; however \texttt{joblib} is very lightweight---less than 0.5 MB pre-built, with no required dependencies.
Thus a more conservative approach that guarantees an exact version is used would also be reasonable in this case.

@@
